Untitled

### 2021-08-27 analysis/paper/paper.Rmd
first draft...?

<!-- add to protocols.io -->
Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system following
starch treatments. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
the starch plates were agitated to keep the starches in suspension in the solutions, 
and during treatments, the rpm was increased to 60.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU five times daily, 
every two hours, for six minutes. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This cycle was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples, that only received
sucrose as a treatment, were included to detect starch contamination.

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were added in triplicates to two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a... 
<!-- photometric reader specs -->
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity.
For a more detailed protocol, see <!-- protocols.io DOI -->

@@

Starch granule counts were divided into three size categories: small (<10 $\mu$m),
medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
potato starches were combined for the mixed treatment samples, as it is not 
possible to distinguish the small starch granules between the two species.

@@

### Extraction from samples

Extraction of starches from the calculus samples was performed by dissolving the
calculus in ethylenediaminetetraacetic acid (EDTA), as it has been shown to
have little to no effect on starch granules (especially native granules)
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using 
the pipette. It was not possible to count all wheat starches
on a slide, so wheat and mix counts were extrapolated as described for the treatment
solutions.

@@

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.

All scripts and data are available on OSF (https://osf.io/uc5qy/)
and Github (https://github.com/bbartholdy/byoc-starch).

@@

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for separate analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (supp. mat).

@@

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m,
small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
total count of small granules within mixed-treatment solutions which are predominantly
wheat granules
(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed-treatment had the highest absolute count of starch granules in solution,
while the biofilms exposed to the wheat solution preserved the greatest number of
granules. The potato treatment had the lowest absolute counts in both the solution 
and in the biofilm samples <!-- insert counts? -->
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

@@

```{r sep-plot, fig.cap="Proportion (%) of sizes of (A) starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

<!-- concentrations of solutions vs. starches in a mouthful of food 
starches in a potato, etc. 
Contact Steph -->

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10 -- 20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
This was especially true for the potato starches, which can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
This is consistent with the findings by Power and collagues [-@powerChimpCalculus2015]
mentioned previously, which found that the larger (defined as around 10 -- 20 $\mu$m) 
granule-producing plants were overrepresented; although, the representation of granules 
larger than 20 $\mu$m in that study is unclear. Since those results were obtained
on a modern population of chimps that were only buried for a short period of time, 
degradation was likely limited [@powerChimpCalculus2015]. The taphonomy of
starch granules has been addressed previously, and depends on burial environment
(pH, temperature, microorganisms), as well as processing of the granules prior to
deposition [@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
The potential for detrimental diagenetic effects on starch granules increases as 
the size and amylose content decreases
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
Combined, the effect of intra-oral starch incorporation
and the bias against large granules, together with the increased effect of taphonomy
on  small granules, would eliminate a large portion of the consumed starch granules,
and would explain why the granules extracted from archaeological dental calculus 
are often sparse.

Additionally, the size of the calculus deposit seems to influence the quantity of 
starches extracted from the calculus, as we found a strong positive correlation
between biofilm size and retained starch granules
(Figure \@ref(fig:cor-plot)), which contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
The lack of a correlation
in archaeological contexts could be due to diagenetic effects; or, the absence of 
amylase activity in our study could be impacting our interpretations.

@@

The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of significant age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].

@@

only briefly retained in the oral cavity. Other granules that are broken off

@@

mineralised plaque would be (at least to some extent) protected from salivary
clearance and mechanical cleaning actions of the tongue and lips. This may explain 
why a previous study found starches more commonly in clusters than dispersed over
the surface of the dental calculus [@powerSEMCalculus2014]. The authors hypothesised 
that the granules are either deposited in clusters, or group together in cracks and

@@

in plaque/calculus will still be susceptible to hydrolysis, as $\alpha$-amylase has

@@

<!-- These channels and pockets of organic matter within the inorganic matrix are used by the 
bacteria for transportation of nutrients, and their persistence in the otherwise 
mineralised matrix may be the result of the differing 
rate of mineralisation displayed by the bacteria [@tanStudyBacterialViability2004].-->

@@

activity is discussed elsewhere<!--Bartholdy et al. in prep-->.
<!--
Starches also have the ability to adhere to certain bacteria [@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
Following deposition and/or mineralisation events, the granules that are able to 
resist hydrolysis and 
mechanical removal will then be trapped in pockets within the mineral matrix, and
will temporarily be protected from shearing action and salivary 
clearance. -->

@@

@@

<!-- limitations of the platform show which questions would be unanswered -->
This article presents preliminary research exploring the potential of an oral 
biofilm model for dietary research in archaeology. A limitation of this study 
is the absence of amylase in the model. The presence of amylase will likely affect 
the total granule counts as well as the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
In future experiments, we recommend the addition of $\alpha$-amylase to the model
in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
<!-- foreshadowing...? -->
While we are able to show the differing 
incorporation caused by absolute counts available, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogenous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
to directly
explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting. The addition of known starch species in
separate samples also removes the identification bias against the smaller
undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring between individuals, and difficulties in accurately documenting 
the entirety of starch granules entering the oral cavity at any given moment,
especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.  
Further research is necessary to explore the mechanisms of the intra-oral starch 
incorporation and retention in dental calculus, and the dental calculus model
presented in this study is uniquely suited to explore these questions and 
may improve interpretations of dietary research in past populations.
<!-- still need to replicate results and do more science, but pretty cool, right? -->

@@

<!-- Forward-looking statements about the field as a whole -->
While we have shown multiple factors that influence the likelihood 
of incorporation, it seems the process is somewhat stochastic, and possibly related 
to the biofilm mineralisation events (as discussed above), which are irregular and
related to various factors. 
The dental calculus biofilm model presented in this study is a powerful tool to 
investigate various aspects of dietary research and starch incorporation and
retention in dental calculus.

>>>>

<!-- add to protocols.io -->
Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system before
starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatment plate, 
the plates were agitated to keep the starches in suspension in the solutions. 
During treatments, the rpm was increased to 60 to facilitate contact between
starch granules and biofilms.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment, were included to detect starch contamination from the
environment, or cross-contamination from other wells in the same plate.

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a... 
<!-- photometric reader specs -->
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see <!-- protocols.io DOI -->

@@

Starch granule counts were divided into three size categories: small (<10 $\mu$m),
medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
potato starches were combined for the mixed-treatment samples, as it was not 
possible to distinguish betweem the small starch granules from the two species.

@@

### Extraction method

Extraction of starches from the calculus samples was performed by dissolving the
calculus in ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. Wheat and mix counts were extrapolated as described above.

@@

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg 
calculus (also z-score standardised) to account for differences in starch 
concentration in the calculus [as per 
@wesolowskiEvaluatingMicrofossilContent2010].

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
All protocols are available on protocols.io.

@@

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

@@

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m,
small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
total count of small granules within mixed-treatment solutions, which are predominantly
wheat granules
(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r subset(count_sol, treatment == "mix")$total`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r subset(count_sample, treatment == "wheat")$total`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r subset(count_sol, treatment == "potato")$total`)
and in the biofilm samples 
(`r subset(count_sample, treatment == "potato")$total`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

@@

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10 -- 20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
This was especially true for the potato starches, which can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
This is consistent with the findings by Power and collagues [-@powerChimpCalculus2015]
mentioned previously, which found that the large (defined as medium in this study) 
granule-producing plants were overrepresented; although, the representation of granules 
larger than 20 $\mu$m in that study is unclear. Since those results were obtained
on a modern population of chimps that were only buried for a short period of time, 
degradation was likely limited [@powerChimpCalculus2015]. The taphonomy of
starch granules has been addressed previously, and depends on burial environment
(pH, temperature, microorganisms), as well as processing of the granules prior to
deposition [@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
The potential for detrimental diagenetic effects on starch granules increases as 
the size and amylose content decreases
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
Combined, the effect of intra-oral starch incorporation
and the bias against large granules, together with the increased effect of taphonomy
on  small granules, would eliminate a large portion of the consumed starch granules,
and could explain the large descrepency between our counts (unaffected by hydrolysis and taphonomy) and the aforementioned studies.  
Another potentially important factor from our results is the size of the calculus
deposit, which seems to influence the quantity of 
starches extracted from the calculus, as we found a strong positive correlation
between biofilm size and retained starch granules
(Figure \@ref(fig:cor-plot)); a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
The lack of a correlation
in archaeological contexts could be due to diagenetic effects; or, the absence of 
amylase activity in our study could be impacting our interpretations.

@@

The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of significant age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  

@@

only briefly present in the oral cavity. Other granules that are broken off

@@

mineralised plaque are (at least to some extent) protected from salivary
clearance and mechanical cleaning actions of the tongue and lips, especially once
a new layer of plaque has covered the surface of the calculus. This may explain 
why a previous study found starches more commonly in clusters, rather than 
dispersed over the surface of the dental calculus [@powerSEMCalculus2014]. 
The authors hypothesised that the granules are either deposited in clusters, or 
group together in cracks and

@@

in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has

@@

@@

activity is discussed elsewhere (Bartholdy et al. in prep.).

@@

<!--Starches also have the ability to adhere to certain bacteria [@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
Following deposition and/or mineralisation events, the granules that are able to 
resist hydrolysis and 
mechanical removal will then be trapped in pockets within the mineral matrix, and
will temporarily be protected from shearing action and salivary 
clearance. -->

@@

<!-- limitations of the platform show which questions would be unanswered -->
This article presents preliminary research exploring the potential of an oral 
biofilm model for dietary research in archaeology. A limitation of this study 
is the absence of $\alpha$-amylase in the model. The presence of $\alpha$-amylase will likely affect 
the total granule counts as well as the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
the lack of $\alpha$-amylase in the system is a beneficial side effect, 
as it can allow us to directly explore
the effect of $\alpha$-amylase on starch counts in future experiments,
where $\alpha$-amylase can be added to the model
in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. <!-- foreshadowing...? -->
While we are able to show the differing 
incorporation caused by absolute counts available, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogenous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting, as many of the variables can be adjusted. 
The addition of known starch species in
separate samples also removes the identification bias against the smaller
undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring between individuals, and difficulties in accurately documenting 
the entirety of starch granules entering the oral cavity at any given moment,
especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.
It can also facilitate training of students and researchers on methods of 
dental calculus analysis, such as starch and phytolith extraction and 
identification, where it can replace the use of finite archaeological resources.  
<!-- we still need to replicate results and do more science... but pretty cool, right? -->

@@

<!-- Forward-looking statements about the field as a whole -->
While we have shown multiple factors that influence the likelihood 
of incorporation, the process still appears to be somewhat stochastic. Further
research is needed to make sense of the contributing factors, and to explore the
mechanisms of intra-oral starch incorporation and retention in dental calculus.
The dental calculus model presented in this study is uniquely suited to explore 
these questions and may improve interpretations of dietary practices in past populations.
